# Reproducible Research: Peer Assessment 1


## Loading and preprocessing the data
```{r echo=TRUE}
activity <- read.csv("activity.csv", header=TRUE)
str(activity)
head(activity)
```


## What is mean total number of steps taken per day?
```{r echo=TRUE}
total_steps_per_day <- tapply(activity$steps, activity$date, sum)
hist(total_steps_per_day, breaks=50, xlab="Steps per day",
     main="Total steps per day")
# interestingly, the mean and median computed by summary() are different from
# those computed by mean() and median()!
summary(total_steps_per_day)
mean(total_steps_per_day, na.rm=T)
median(total_steps_per_day, na.rm=T)
```

## What is the average daily activity pattern?
Plot the 5-minute intervals vs the mean steps taken per interval.
```{r echo=TRUE}
interval_groups <- split(activity, activity$interval)
mean_steps_per_interval <- sapply(interval_groups,
                                  function(x) {mean(x$steps, na.rm=T)})
plot(names(mean_steps_per_interval), mean_steps_per_interval, type="l",
     main="Average steps per 5 minute time interval",
     xlab="5 minute interval", ylab="Mean steps per interval")
```

Determine which 5-minute interval contains the maximum number of steps.
```{r echo=TRUE}
total_steps_per_interval <- sapply(interval_groups,
                                   function(x) {sum(x$steps, na.rm=T)})
which.max(total_steps_per_interval)
# the 835 interval contains the max number of steps; 104 is just the index of the
# 835 entry in the vector
```

## Imputing missing values
The total number of missing values in the dataset:
```{r echo=TRUE}
num_missing_values <- dim(activity)[1] - sum(complete.cases(activity))
num_missing_values
```

Impute missing values using the mean for the 5-minute interval in question.  And
create a new instance of the dataset containing the imputed data.
```{r echo=TRUE}
good <- complete.cases(activity)
get_interval_mean <- function(interval) {
  interval_ch <- as.character(interval)
  idx <- which(names(mean_steps_per_interval)==interval_ch)
  mean_steps_per_interval[idx]
}
activity2 <- activity
activity2$steps[!good] <- sapply(activity2$interval[!good], FUN=get_interval_mean)
# verify the missing values are all gone
num_missing_values2 <- dim(activity2)[1] - sum(complete.cases(activity2))
num_missing_values2
```

Histogram, mean and median of total steps per day.
```{r echo=TRUE}
total_steps_per_day2 <- tapply(activity2$steps, activity$date, sum)
hist(total_steps_per_day2, breaks=50, main="Histogram of Total Steps Per Day",
     xlab="Steps")
summary(total_steps_per_day2)
mean(total_steps_per_day2, na.rm=T)
median(total_steps_per_day2, na.rm=T)
```

## Are there differences in activity patterns between weekdays and weekends?

Create a new factor variable indicating the type of day: weekday or weekend
```{r echo=TRUE}
day_of_week <- weekdays(as.Date(activity2$date))
get_day_type <- function(x){
  if (x %in% c("Saturday", "Sunday"))
    return("weekend")
  else
    return("weekday")
}
day_type <- sapply(day_of_week, get_day_type)
day_type <- as.factor(day_type)
table(day_type)
```

Plot the 5-minute intervals vs mean steps taken per interval for each of the two day type groups: weekday and weekend
```{r echo=TRUE}
day_type_group <- split(activity2, day_type)
interval_groups <- split(day_type_group$weekday, day_type_group$weekday$interval)
mean_steps_per_interval_weekday <- sapply(interval_groups,
                                          function(x) {mean(x$steps, na.rm=T)})
interval_groups <- split(day_type_group$weekend, day_type_group$weekend$interval)
mean_steps_per_interval_weekend <- sapply(interval_groups,
                                          function(x) {mean(x$steps, na.rm=T)})

par(mfrow=c(2,1))
plot(names(mean_steps_per_interval_weekend), mean_steps_per_interval_weekend,
     type="l",
     main="Average steps per 5 minute interval - weekend",
     xlab="5 minute interval", ylab="Mean steps per interval")
plot(names(mean_steps_per_interval_weekday), mean_steps_per_interval_weekday,
     type="l",
     main="Average steps per 5 minute interval - weekday",
     xlab="5 minute interval", ylab="Mean steps per interval")
par(mfrow=c(1,1))
```









